Faithful Variable Screening for High-Dimensional Convex Regression

### 2014-09-11 tex/scam/optimization.tex
modified the order statistic notation

\def\uds#1{#1}

@@

Let $\bds{x}_{i}\in\mathbb{R}^{p}$ be the covariate, let $y_{i}$ be
the response and let $\epsilon_{i}$ be the mean zero noise. The
regression function $f(\cdot)$ we estimate is the sum of
univariate functions $f_{k}(\cdot)$ in each variable dimension and a scalar
offset $\mu$.  We impose additional constraints that each
function $f_{k}(\cdot)$ is convex function, which can be
represented by its supporting hyperplanes, i.e.,
\begin{equation}\label{hyper}
      f_{i'k} \geq f_{ik} + \beta_{ik}(x_{i'k}-x_{ik}) \quad
      \textrm{for all $i,i' = 1,\ldots, n$,}
\end{equation}
where $f_{ik}\coloneqq f_{k}(x_{ik})$ and $\beta_{ik}$ is the
subgradient at point $x_{ik}$. This apparently requires $O(n^2 p)$ constraints to
impose the supporting hyperplane constraints.
In fact, only $O(np)$
constraints suffice, since univariate convex functions are
characterized by the condition that the subgradient, which is a scalar, must
increase monotonically. This observation leads to the  optimization
\begin{equation}
\begin{split}
       \min_{f,\beta,\mu} & \;\; \frac{1}{2n}\sum_{i=1}^{n}
                     \Bigl( y_{i}-\mu - \sum_{k=1}^{p}f_{ik}\Bigr)^{2} 
                         + \lambda\sum_{k=1}^{p}\|f_k\|_{\infty} \\
       \textrm{subject to} &\;\; \textrm{for all $k=1,\ldots, p$:}\\
       & \;\; f_{(i+1)_k k} = f_{(i)_k k} +
       \beta_{(i)_k k}(x_{(i+1)_k k}-x_{(i)_k k}),\;\textrm{for $i=1,\ldots, n-1$}\\
       & \;\; \sum_{i=1}^{n}f_{ik}=0,\\
       & \;\; \beta_{(i+1)_k k} \geq \beta_{(i)_k k}\;\textrm{for $i=1,\ldots, n-1$}.
\end{split}
\label{np}
\end{equation}
Here $f_k$ denotes the vector $f_k = (f_{1k}, f_{2_k}\ldots, f_{nk})^T\in\reals^n$
and $\{(1)_k,(2)_k,\ldots,(n)_k\}$ are the indices in the sorted ordering
of the values of coordinate $k$:
\begin{equation}
x_{(1)_k k} \leq{} x_{(2)_k k} \leq \cdots \leq{} x_{(n)_k k}.
\end{equation}

We can solve for $\mu$ explicitly as  
$\mu = \frac{1}{n} \sum_{i=1}^n y_i = \bar{y}$.  This follows from the
KKT conditions
and the constraints $\sum_i f_{ki} = 0$.
It is easy to verify that the constraints in (\ref{np}) satisfy the
supporting hyperplane constraints, since
for all $j > i$
\begin{align*}
  f_{(j)_kk}-f_{(i)_kk}  & = \sum\limits_{t=i}^{j-1}(f_{(t+1)_kk}-f_{(t)_kk}) \\
   &= \sum\limits_{t=i}^{j-1}\beta_{(t)_kk}(x_{(t+1)_kk}-x_{(t)_kk}) \\
   &\geq \beta_{(i)_kk}\sum\limits_{t=i}^{j-1}(x_{(t+1)_kk}-x_{(t)_kk}) \\
  & = \beta_{(i)_kk}(x_{(j)_kk}-x_{(i)_kk}) 
\end{align*}
and for all $j < i$
\begin{align*}
f_{k(j)}-f_{k(i)} & =
    \sum\limits_{t=j}^{i-1}(f_{k(t)}-f_{k(t+1)}) \\
     & = \sum\limits_{t=j}^{i-1}\beta_{k(t)}(x_{k(t)}-x_{k(t+1)}) \\
     &\geq \beta_{k(i)}\sum\limits_{t=j}^{i-1}(x_{k(t)}-x_{k(t+1)}) \\
     & = \beta_{k(i)}(x_{k(j)}-x_{k(i)}).
\end{align*}

@@

The sparse convex additive model optimization in (\ref{np}) is a quadratic program with
$O(np)$ variables and $O(np)$ constraints. 
Directly applying a QP solver for $f$ and $\beta$
is computationally expensive for relatively large
$n$ and $p$. However, notice that variables in different feature
dimensions are only coupled in the squared error term
$(y_{i}-\mu - \sum_{k=1}^{p}f_{ik})^{2}$. Hence, we can apply the block
coordinate descent method, where in each step we solve the following
QP subproblem for $\{f_k, \beta_k\}$ with the
other variables fixed. In matrix notation, the optimization is
\begin{align}
\begin{split}
\min_{ f_k, \beta_k, \gamma_k} \;\;& \frac{1}{2n} \| \uds{r}_k - \uds{f}_k \|_2^2 
     + \lambda \gamma_k \label{opt:1d_compact} \\
 \textrm{such that } & P \uds{f}_k = \diag(P \bds{x}_k)  \uds{\beta}_k \\
   & D \uds{\beta}_k \leq 0 \\
   & -\gamma_k \mathbf{1}_n \leq \uds{f}_k \leq \gamma_k \mathbf{1}_n   \\
   & \mathbf{1}_n^\tran \uds{f}_k = 0 
\end{split}
\end{align}
where $\uds{\beta}_k \in \R^{n-1}$ is the vector $\uds{\beta}_k =
(\beta_{1k}, \ldots, \beta_{(n-1)k})^T$, and
$\uds{r}_{k} \in \R^n$ is the residual vector $\uds{r}_{k} = (y_i -
\hat\mu - \sum_{k' \neq k} f_{ik'})^T$.
In addition, 
$P \in \R^{(n-1) \times n}$ is a permutation matrix where the $i$-th
row  is all zeros except for a $-1$ in position $(i)_k$ and a $1$ in
position $(i+1)_k$, and $D \in \R^{(n-2) \times (n-1)}$ is another
permutation matrix  where the $i$-th row is all zeros except for a $1$ 
in position $i$ and a $-1$ in position $i+1$.  We denote by
$\diag( v )$ the diagonal matrix with diagonal entries $v$.
The extra variable $\gamma_{k}$ is introduced to impose the
regularization penalty involving the $\ell_{\infty}$ norm.

@@

The univariate concave function estimation is a straightforward
modification of optimization~\ref{opt:1d_compact}. It is only
necessary to modify the linear inequality constraints so that the subgradients are
non-increasing: $\beta_{(i+1)_kk} \leq \beta_{(i)_kk}$.

>>>>

\def\uds#1{#1}
\def\perm#1{\pi_k(#1)}

@@

Let $\bds{x}_{i}\in\mathbb{R}^{p}$ be the covariate, let $y_{i}$ be
the response and let $\epsilon_{i}$ be the mean zero noise. The
regression function $f(\cdot)$ we estimate is the sum of
univariate functions $f_{k}(\cdot)$ in each variable dimension and a scalar
offset $\mu$.  We impose additional constraints that each
function $f_{k}(\cdot)$ is convex, which can be
represented by its supporting hyperplanes, i.e.,
\begin{equation}\label{hyper}
      f_{i'k} \geq f_{ik} + \beta_{ik}(x_{i'k}-x_{ik}) \quad
      \textrm{for all $i,i' = 1,\ldots, n$,}
\end{equation}
where $f_{ik}\coloneqq f_{k}(x_{ik})$ is the function value and $\beta_{ik}$ is a
subgradient at point $x_{ik}$. This ostensibly requires $O(n^2 p)$ constraints to
impose the supporting hyperplane constraints.
In fact, only $O(np)$
constraints suffice, since univariate convex functions are
characterized by the condition that the subgradient, which is a scalar, must
increase monotonically. This observation leads to the  optimization
\begin{equation}
\begin{split}
       \min_{f,\beta,\mu} & \;\; \frac{1}{2n}\sum_{i=1}^{n}
                     \Bigl( y_{i}-\mu - \sum_{k=1}^{p}f_{ik}\Bigr)^{2} 
                         + \lambda\sum_{k=1}^{p}\|f_k\|_{\infty} \\
       \textrm{subject to} &\;\; \textrm{for all $k=1,\ldots, p$:}\\
       & \;\; f_{\perm{i+1} k} = f_{\perm{i} k} +
       \beta_{\perm{i} k}(x_{\perm{i+1} k}-x_{\perm{i} k}),\;\textrm{for $i=1,\ldots, n-1$}\\
       & \;\; \sum_{i=1}^{n}f_{ik}=0,\\
       & \;\; \beta_{\perm{i+1} k} \geq \beta_{\perm{i} k}\;\textrm{for $i=1,\ldots, n-1$}.
\end{split}
\label{np}
\end{equation}
Here $f_k$ denotes the vector $f_k = (f_{1k}, f_{2_k}\ldots, f_{nk})^T\in\reals^n$
and $\{\perm{1},\perm{2},\ldots,\perm{n}\}$ are the indices in the sorted ordering
of the values of coordinate $k$:
\begin{equation}
x_{\perm{1} k} \leq{} x_{\perm{2} k} \leq \cdots \leq{} x_{\perm{n} k}.
\end{equation}

We can solve for $\mu$ explicitly as  
$\mu = \frac{1}{n} \sum_{i=1}^n y_i = \bar{y}$.  This follows from the
KKT conditions
and the constraints $\sum_i f_{ki} = 0$.
It is easy to verify that the constraints in (\ref{np}) satisfy the
supporting hyperplane constraints, since
for all $j > i$
\begin{align*}
  f_{\perm{j}k}-f_{\perm{i}k}  & = \sum\limits_{t=i}^{j-1}(f_{\perm{t+1}k}-f_{\perm{t}k}) \\
   &= \sum\limits_{t=i}^{j-1}\beta_{\perm{t}k}(x_{\perm{t+1}k}-x_{\perm{t}k}) \\
   &\geq \beta_{\perm{i}k}\sum\limits_{t=i}^{j-1}(x_{\perm{t+1}k}-x_{\perm{t}k}) \\
  & = \beta_{\perm{i}k}(x_{\perm{j}k}-x_{\perm{i}k}) 
\end{align*}
and for all $j < i$
\begin{align*}
f_{k(j)}-f_{k(i)} & =
    \sum\limits_{t=j}^{i-1}(f_{k(t)}-f_{k(t+1)}) \\
     & = \sum\limits_{t=j}^{i-1}\beta_{k(t)}(x_{k(t)}-x_{k(t+1)}) \\
     &\geq \beta_{k(i)}\sum\limits_{t=j}^{i-1}(x_{k(t)}-x_{k(t+1)}) \\
     & = \beta_{k(i)}(x_{k(j)}-x_{k(i)}).
\end{align*}

@@

The sparse convex additive model optimization in (\ref{np}) is a quadratic program with
$O(np)$ variables and $O(np)$ constraints. 
Directly applying a QP solver for $f$ and $\beta$
is computationally expensive for relatively large
$n$ and $p$. However, notice that variables in different feature
dimensions are only coupled in the squared error term
$(y_{i}-\mu - \sum_{k=1}^{p}f_{ik})^{2}$. Hence, we can apply the block
coordinate descent method, where in each step we solve the following
QP subproblem for $\{f_k, \beta_k\}$ with the
other variables fixed. In matrix notation, the optimization is
\begin{align}
\begin{split}
\min_{ f_k, \beta_k, \gamma_k} \;\;& \frac{1}{2n} \| \uds{r}_k - \uds{f}_k \|_2^2 
     + \lambda \gamma_k \label{opt:1d_compact} \\
 \textrm{such that } & P \uds{f}_k = \diag(P \bds{x}_k)  \uds{\beta}_k \\
   & D \uds{\beta}_k \leq 0 \\
   & -\gamma_k \mathbf{1}_n \leq \uds{f}_k \leq \gamma_k \mathbf{1}_n   \\
   & \mathbf{1}_n^\tran \uds{f}_k = 0 
\end{split}
\end{align}
where $\uds{\beta}_k \in \R^{n-1}$ is the vector $\uds{\beta}_k =
(\beta_{1k}, \ldots, \beta_{(n-1)k})^T$, and
$\uds{r}_{k} \in \R^n$ is the residual vector $\uds{r}_{k} = (y_i -
\hat\mu - \sum_{k' \neq k} f_{ik'})^T$.
In addition, 
$P \in \R^{(n-1) \times n}$ is a permutation matrix where the $i$-th
row  is all zeros except for a $-1$ in position $\perm{i}$ and a $1$ in
position $\perm{i+1}$, and $D \in \R^{(n-2) \times (n-1)}$ is another
permutation matrix  where the $i$-th row is all zeros except for a $1$ 
in position $i$ and a $-1$ in position $i+1$.  We denote by
$\diag( v )$ the diagonal matrix with diagonal entries $v$.
The extra variable $\gamma_{k}$ is introduced to impose the
regularization penalty involving the $\ell_{\infty}$ norm.

@@

The univariate concave function estimation is a straightforward
modification of optimization~\ref{opt:1d_compact}. It is only
necessary to modify the linear inequality constraints so that the subgradients are
non-increasing: $\beta_{\perm{i+1}k} \leq \beta_{\perm{i}k}$.
